Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan

### 2019-11-21 analysis/paper/paper.Rmd
Change some terms and clarify sentenses

Kiwulan (Figure \@ref(fig:KWL-map)) was a rescue archaeology project that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge. The total area includes eight open area sections of 262 excavation squares (4 m by 4 m) reaching 3,814 m^2^ in total [@Chen2007]. The archaeological evidence includes a large amount of artefacts, burials, middens, post holes, wooden pillars, and stone structures, all of which indicates it was a settlement. Most ornaments were found in situ while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The ornaments studied in this paper were found from 40 adjacent squares in the largest open area, located in the middle part of the excavation, see Figure \@ref(fig:KWL-sam-area). They were sampled because those units were stratigraphically intact with the least disturbance by modern construction activity, compared to excavation squares on the periphery of the site.

The chronology of Kiwulan can be divided into two phases represented by a Lower Layer Culture (700-1200 AD) and a Upper Layer Culture (1400-1900 AD) separated by a sterile. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates previously published by @Chen2007. This paper focuses on the Upper Layer Culture because this has a sequence spanning pre-European contact, European presence, and Chinese presence. Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. Due to ambiguities in the previous chronology, I re-examined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European period, European period, and Chinese period to indicate the dominant foreign influence and cultural interaction for different time periods.

@@

This study focuses on `r sam_post_hole` ornaments from 40 sampling squares that were excavated _in situ_ from the household area of Kiwulan, indicated by numerous post-holes (Figure \@ref(fig:KWL-sam-area)). Thirty burials contained 3,173 ornaments, and 27 ornaments were recovered from middens. In the burials ornaments were in the form of bead strings that consisted of hundreds to thousands of pieces per burial. The ornaments from the burial contexts are not analysed here since few burials can be confidently assigned to a time period. The household area represents domestic spaces suitable for testing for changes in ornament distribution relating to social organisation. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament are golden beads (n = `r golden_bead_n`), followed by glass beads (n =  `r glass_bead_n`), agate beads (n = `r agate_bead_n`), metal rings (n = `r metal_ring_n`), and metal bells (n = `r metal_bell_n`, see also Figure \@ref(fig:ornaments-image)). The variety of subtypes and their individual frequencies are summarized in Table \@ref(tab:tidy-data-for-table).

>>>>

Kiwulan (Figure \@ref(fig:KWL-map)) was a rescue archaeology project that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge. The total area includes eight open area sections of 262 excavation squares (4 m by 4 m) reaching 3,814 m^2^ in total [@Chen2007]. The archaeological evidence includes a large amount of artefacts, burials, middens, post holes, wooden pillars, and stone structures, all of which indicates it was a settlement. Most ornaments were found while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The ornaments studied in this paper were found from 40 adjacent squares in the largest open area, located in the middle part of the excavation, see Figure \@ref(fig:KWL-sam-area). They were sampled because those units were stratigraphically intact and undisturbed by modern construction activity, compared to excavation squares on the periphery of the site.

The chronology of Kiwulan can be divided into two phases represented by a upper component (700-1200 AD) and a lower component (1400-1900 AD) separated by a sterile. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates previously published by @Chen2007. This paper focuses on the upper component because this has a sequence spanning pre-European contact, European presence, and Chinese presence. Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. Due to ambiguities in the previous chronology, I re-examined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European period, European period, and Chinese period to indicate the dominant foreign influence and cultural interaction for different time periods.

@@

This study focuses on `r sam_post_hole` ornaments from 40 sampling squares that were excavated _in situ_ from the household area of Kiwulan, indicated by numerous post-holes (Figure \@ref(fig:KWL-sam-area)). Thirty burials contained 3,173 ornaments, and 27 ornaments were recovered from middens. In the burials ornaments were in the form of bead strings that consisted of hundreds to thousands of pieces per burial. The ornaments from the burial contexts are not statistically analysed because most burials were assigned to 17-18 th century that offers limited quantitative comparison with other time periods. But the use of ornaments as body adornments in burials is discussed to understand the functions of ornaments. The household area represents domestic spaces suitable for testing for changes in ornament distribution relating to social organisation. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament are golden beads (n = `r golden_bead_n`), followed by glass beads (n =  `r glass_bead_n`), agate beads (n = `r agate_bead_n`), metal rings (n = `r metal_ring_n`), and metal bells (n = `r metal_bell_n`, see also Figure \@ref(fig:ornaments-image)). The variety of subtypes and their individual frequencies are summarized in Table \@ref(tab:tidy-data-for-table).
